Class differences in social networks: Evidence from a referral experiment

### 2025-06-24 writing/manuscript.tex
v1 chapter 3

\usepackage{blindtext}

\usepackage[colorlinks=true, citecolor=blue, linkcolor=blue, urlcolor=blue,breaklinks]{hyperref}

@@

\section*{Abstract}
Lorem Ipsum \citep{beaman_job_2018}
\medskip \\
\textbf{JEL Classification:} C93, D03, D83, J24 \\
\textbf{Keywords:} productivity beliefs, referrals, field experiment, skill identification, social class

% Problem: Labor markets increasingly value both cognitive and social skills, but firms struggle to identify these skills in candidates. Referrals are a common hiring tool, but their effectiveness across different skill types and potential social class biases remain unclear.
% Method: Using a field experiment with 849 university students, we study how well peers identify cognitive and social skills of their classmates.
% Key Results: Students effectively identify cognitive skills but struggle with social skills. Limited evidence of social class bias in referrals.
% Implications: Referrals may be more effective for identifying certain types of skills than others.
% We study how well referrals help screening for the two types of worker skills within the same network and compare the differential skill identification ability of referrers.

\pagebreak
\section{Introduction}
Equally qualified individuals face different labor market outcomes depending on their socioeconomic status \citep{stansbury2024class}. A key driver of this inequality is due to differences in social capital.\footnote{See for example \citet{loury_dynamic_1977,bourdieu_forms_1986} for pioneering work on the relationship between social position and human capital acquisition.} Because it correlates strongly with labor market income, the most important facet of social capital is the share of high-SES connections among low-SES individuals \citep{chetty2022na}. A lack of social capital means lack of access to individuals with influential (higher paid) jobs and job opportunities. In economic terms, it implies having worse outcomes when using one's network to find jobs conditional on the capacity on leveraging one's social network.\footnote{See for example \citet{lin1981asr,mouw2003asr} for differential outcomes while using contacts in job search, and \citet{smith2005ajs,Pedulla2019RaceAN} specifically for the effects of race conditional on network use.}

Referral hiring, the formal or informal process where firms ask workers to recommend qualified candidates for job opportunities, is a common labor market practice which makes evident the role of differences in social capital. As referrals originate from the networks of referrers, the composition of referrer networks becomes a crucial channel that propagates inequality: Similar individuals across socio-demographic characteristics form connections at higher rates \citep{mcpherson2001birds}, making across SES (low-to-high) connections less likely than same-SES connections \citep{chetty2022na}. Referrals will thus reflect similarities in socio-demographic characteristics present in networks even in the absence of biases in the referral procedure, i.e., referring at random from one's network according to some productivity criteria.

Yet, experimental evidence shows referrals can be biased even under substantial pay-for-performance incentives beyond what is attribuable to differences in network compositions, at least for the case of gender \citep{beaman_job_2018,hederos2025gender}. A similar bias against low-SES may further exacerbate outcomes of low-SES individuals: If job information are in the hands of a select few high-SES which low-SES have already limited network access to (social capital hypothesis), and high-SES referrers are biased against low-SES, referring other high-SES at higher rates than their network composition, we should expect referral hiring to further disadvantagee low-SES.

The empirical question we answer in this paper is whether referrers are biased against low-SES peers after accounting for differences in the network SES compositon. We also evaluate the causal impact of two different incentive structures on referral behavior.

In this study, we study inequalities related to SES combining a university-wide cross-sectional network data set comprising over 4,500 students in which classroom interactions are recorded along with individual attributes. We focus on the role of SES in referrals by experimentally investigating whether individuals who are asked to refer a peer tend to refer a same-SES candidate. We also explore potential mechanisms behind referral patterns by randomizing participants into two different incentive structures. To this end, we conducted a lab-in-the-field experiment with 734 students in a Colombian university. Participants were instructed to refer a qualified student for tasks similar to the math and reading parts of the national university entry exam (equivalent of SAT in US system). To incentivize participants to refer qualified candidates, we set earnings dependent on referred candidates' actual university entry exam scores.

Referral hiring in the labor market can range from firm-level formal referral programs asking employees to bring candidates to simply passing on job opportunities between network members \citep{topa2019social}. As our participants are students at the university and refer based on exam scores, we abstract away from formal referral programs with defined job openings. Our setting instead resembles situations where contacts share opportunities with each other without the need for the referred candidate to take any action and without revealing the identity of the referrer. This eliminates reputational concerns as there is no hiring firm, and puts a lower bound on the expected reciprocity for the referrer in combination with pay-for-performance incentives \citep{witte2021workers,bandiera2009social}. At the same time, referring based on university entry exam scores are still an objective, widely accepted measure of ability, and we show evidence that referrers in our setting not only possess accurate information about these signals but are also able to screen more productive individuals from their university network.

In a university setting, class attendance provides essential opportunities for face-to-face interaction between students. On the one hand, this reduces network segregation by providing ample opportunities to meet across-SES, because of the exposure to an equal or higher level of high-SES compared to the population \citep{chetty2022n}.\footnote{In a different sample from the same university population, \citet{diaz_peer_2025} show this holds true for the highest-SES individuals at this institution, accounting for about 6\% of their sample but less than 5\% of Colombian high-school graduates \citet{fergusson2021desigualdad}.} On the other hand, as students take more and more classes together, their similarities across all observable characteristics tend to increase \citep{kossinets_origins_2009}, which should drive the high- and low-SES networks to segregate. Our setting is ideal to study these opposing forces: First, The very high level of income inequality and existence of deeply rooted historical groups in Colombia makes SES differences extremely visible in access to tertiary education, where the rich and poor typically select into different institutions \citep{jaramillo-echeverri2023}. Yet, thanks to the particular standing of the institution we have chosen for this study (Figure \ref{school_quality_income}), all SES groups including both low- and high-SES mix together in this university. Second, using administrative data, we are able to reconstruct 734 participants' complete university network based on the number of common courses they have taken together with other students. This allows directly identifying the individual characteristics of those getting referrals among all possible candidates, as well as descriptive characterizations of similarity (e.g., in same-SES share) in student networks as a function of the number of classes taken.

We find strong evidence that networks of high- and low-SES participants exhibit same-SES bias. Both groups are connected at higher rates with their own SES group than what would be at random given actual group shares at the university (Figure \ref{availability_lses}). As students take more courses together within the same program, their networks dwindle in size (Figures \ref{share_same_program_by_tie} and \ref{conns_by_tie}), and become more homogenous in SES-shares (Figure \ref{share_same_ses_by_tie}). We identify selection into academic programs as a key mechanism. The private university where our study took place implements exogenous cost-based program pricing and does not offer SES-based price reductions. These result in programs with very large cost differences within the same university (Figure \ref{fees_with_boxplot}). We find that average yearly fee paid per student increases with SES, and the high-SES share in the most expensive program at the university, medicine, drives the network segregation across SES (Figure \ref{ses_distribution_by_fees}).

Do segregated networks account for all the differences in SES referral rates across SES groups? Although same-SES referrals are 17\% more common than is suggested by referrer networks, controlling for these, we find no general SES-bias against beyond what is attribuable to network composition. Regardless of SES, participants refer productive individuals, and referred candidates are characterized by a very high number of courses taken together. The latter underlies the impact of program selection, where smaller and more homogenous parts of the networks are activated for referrals made in our setting. Our treatment randomized participants across two different incentive schemes by adding a substantial monetary bonus (\$25) for the referred candidate on top of the pay-for-performance incentives. We provide evidence that treatment incentives did not change the referral behavior across the same-SES referral rate, the number of courses taken together with the referral candidate, and the candidate's exam scores.

This paper contributes to the literature on referral experiments by solving the challenge of observing the entire referral network. Earlier research could only compare referrals made across different incentive structures or experimental instructions and make according conclusions. For example, when participants are paid on the basis of their referred candidate's productivity instead of receiving a fixed finder's fee \citep{beaman_who_2012}, or when participants are restrictred to refer either a male or female candidate instead of freely \citep{beaman_job_2018}. \citet{pallais_why_2016} recruited a random sample of nonreferred workers to compare with referred ones, but none of the previous studies could provide a direct comparison of the referral choice set with those who were selected by participants. Closest to our work is the work of \citet{hederos2025gender}, who elicited friendship networks by asking referrers to name 5 friends. Their findings suggest only half of those who were referred were from the elicited friendship network, and thus is not a complete observation of the referral choice set. Although commonplace, censored elicitation methods also result in underestimating network effects \citep{griffith2022jle} and may suffer from biases in recall. We are able to take our analysis one step further by asking for referrals from the enrollment network, where we have complete information on every single connection that may or may not get a referral. This allows us to neatly separate the effect of the network composition from any potential biases stemming from the referral procedure itself.

Second, we build upon to the earlier work on SES-biases in referrals. To our knowledge, the first to study SES-biases in referals are \citet{diaz_peer_2025}, and our study is conceptually the closest to theirs. Drawing from a similar sample from the same institution, \citet{diaz_peer_2025} focus on referrals from first year students made within mixed-program classrooms, and find no evidence for an aggregate bias against low-SES. We also find no aggregate bias against low-SES in referrals. Our setup differs as we sample from students who completed their first year and impose no limits on referring from a classroom. This has several implications: We find that referrals in our setup go to individuals within the same program, and that programs have different SES-shares which become more even more accentuated as students take more courses together. While networks drive inequality in referral outcomes because of the institional environment in our sample, we have no reason to believe first year student networks in \citet{diaz_peer_2025} have similar levels of segregation to begin with. Following the recent evidence, implementing more mixed-program courses which allow for across-SES mixing can be a clear policy goal \citep{alan2023troeasa,rohrer2021po}.

Finally, we contribute to the growing literature on SES differences in the labor market, expliciting the role of networks as a driver of inequality.
\citet{stansbury2024class} find that low-SES researchers coauthor more often with other low-SES, and have networks that have lower values which can explain why

The remainder of the paper is organized as follows. Section \ref{s2} begins with the background and setting in Colombia. In Section \ref{s3} we present the design of the experiment. In Section \ref{s4} we describe the data and procedures. Section \ref{s5} discusses the results of the experiment. Section \ref{s6} concludes. The Appendix presents additional tables and figures as well as the experiment instructions.

\begin{figure}[H]
  \centering
  \caption{Income, performance, and university choice in Colombia}\label{school_quality_income}
  \includegraphics[width=0.6\linewidth]{../figures/school_quality_income.png}
  \begin{tablenotes}
    \footnotesize
    \item[] \textit{Note:} This figure shows the average score national university entry exam by family income and type of higher education instition. With average student score in the 65-70 band, the private university where we conducted this experiment caters to low- and high-SES students. Figure reproduced from \citet{fergusson2021distincion}.
  \end{tablenotes}
\end{figure}

@@

Our study takes place at UNAB, a medium-sized private university in Bucaramanga, Colombia with approximately 6,000 enrolled students. The university's student body is remarkably diverse with about 35\% of the students classified as low-SES, and 15\% high-SES. Diversity at this institution provides a unique research setting as Colombian society is highly unequal and generally characterized by limited interaction between social classes, with different socioeconomic groups separated by education and geographic residence.\footnote{Colombia has consistently ranked as one of the most unequal countries in Latin America \citep{barretoherrera2024regional}, with the richest decile earning 50 times more than the poorest decile \citep{comisioneconomicaparaamericalatinayel2023}. This economic disparity is reflected by a highly stratified society with significant class inequalities and limited class mobility \citep{angulo2012movilidad,garcia2015loteria}.} Despite significant financial barriers, many lower and middle-SES families prioritize university education for their children \citep[p. 103]{hudson_colombia_2010}, and UNAB represents one of the few environments in Colombia where sustained inter-SES contact occurs naturally (see Figure \ref{school_quality_income}).

In 1994, Colombia introduced a nationwide classification system dividing the population into 6 strata based on housing characteristics and neighborhood amenities.\footnote{Initially designed for utility subsidies from higher strata (5 and 6) to support lower strata (1 to 3), it now extends to university fees and social program eligibility. Stratum 4 neither receives subsidies nor pays extra taxes. This stratification system largely aligns with and potentially reinforces existing social class divisions \citep{uribe2008estratificacion, guevara2019spatializing}.} We use this classification as the measure of SES in our experiment: Students in strata 1 to 2 are categorized as low-SES, strata 3 to 4 as middle-SES and those in strata 5 to 6 as high-SES.

We invited via email all 4,417 UNAB undergraduate students who had at the time of recruitment completed their first year at the university to participate in our experiment. 837 students who joined (19\%) vary in terms of their academic programs, SES, and progress in their studies. This setup provides a unique opportunity for collaborative inter-class contact on equal status, whose positive effects on reducing discrimination are casually documented \citep{rao2019familiarity, mousa2020building,lowe2021types}.

Undergraduate programs at UNAB are spread across two semesters, with each individual course lasting one semester. Students take between 5 to 7 courses per semester, with programs lasting anywhere between 4 to 12 semesters (2 to 6 years). Medicine, the largest program by size at UNAB, lasts for 12 semesters, followed by engineering programs at 10 semesters. Most remaining programs lasting for about 8 to 10 semesters, with specialized programs for immediate entry into the workforce lasting only 4.

@@

\subsection{Productivity measures}
To establish an objective basis for referral productivity, we use national university entry exam scores (SABER 11). These scores provide pre-existing, comparable measures of ability across two domains relevant for the labor market. By using existing administrative data, we eliminate the need for additional testing and ensure that all eligible students have comparable productivity measures. The scores we use in this experiment comprise of critical reading and mathematics parts.

@@

\subsection{Referral task}
After eliciting beliefs about their own performance, participants engage in incentivized peer recommendations. For both test areas (critical reading and mathematics), participants recommend one peer they believe excels in that domain. We first present an example question from the relevant test area to clarify what skills are being assessed. Participants then type the name of their recommended peer, with the system providing autocomplete suggestions from enrolled students who have taken the test (see Figure \ref{interface}).

@@

Participants can only recommend students with whom they have taken at least one class during their university studies. This requirement ensures that referrals are based on actual peer interactions and overlap with the enrollment network that we construct. The order in which participants make recommendations across the two areas is randomized.

We incentivize referrals using a productivity-based payment scheme. Referrers earn increasing monetary rewards as the percentile ranking of their recommendation increases (see Figure \ref{piece_rate_by_score}). We multiply the piece rate coefficient associated to the percentile rank with the actual test scores of the recommendation to calculate earnings. This payment structure provides strong incentives to screen for highly ranked peers, with potential earnings up to \$60 per recommendation.\footnote{Due to the selection into the university, the actual test score distribution has limited variance. Below a certain threshold students cannot qualify for the institution and choose a lower ranked university, and above a certain threshold they have better options to choose from.}

@@

\subsection{Treatment variation}
We implement a between-subjects treatment that varies whether the recommended peer also receives payment. In the \textbf{Baseline} treatment, only the referrer can earn money based on their recommendation's productivity. The \textbf{Bonus} treatment adds an additional fixed payment of \$25 to any peer who is recommended in the randomly selected area for payment. This payment is independent of the peer's actual productivity (see Figure \ref{tab:treatments}).

@@

\subsection{Belief elicitation}
We elicit incentivized beliefs at two points in the experiment. First, before making referrals, participants report their beliefs about their own percentile ranking in each test area. Second, after making each referral, participants report their beliefs about their recommendedation's percentile ranking. For both belief elicitation tasks, participants earn \$5 if their guess is within 7 percentiles of the true value. This tolerance level is expected to balance precision with the difficulty of the task.

@@

\section{Sample, Incentives, and Procedure} \label{s4}
We invited all 4,417 UNAB undergraduate students who had at the time of recruitment completed their first year at the university to participate in our experiment. A total of 837 students took part in the data collection with a 19\% response rate. Our final sample consists of 734 individuals who referred peers with whom they have taken at least one class together, resulting in an 88\% success rate for the sample. We randomly allocated half of the participants into either \textbf{Baseline} or \textbf{Bonus} treatments. Table \ref{tab:balance} presents key demographic characteristics and academic performance indicators across treatments (see Appendix Table \ref{tab:selection} for selection). The sample is well-balanced between the \textbf{Baseline} and \textbf{Bonus} conditions and we observe no statistically significant differences in any of the reported variables (all \textit{p} values $> 0.1$). Our sample is characterized by a majority of middle-SES students with about one-tenth of the sample being high-SES students. The test scores and GPA distributions are balanced. On average, participants took 3.8 courses together with their network, and the average network consisted of 175 peers.

@@

The experiment was conducted online through Qualtrics, with participants recruited from active UNAB students. To manage budget constraints while maintaining sufficient incentives, we randomly selected one in ten participants for payment. Selected participants received a fixed payment of \$17 for completion, plus potential earnings from one randomly selected belief question (up to \$5) and one randomly selected recommendation question (up to \$60), for maximum total earnings of \$82. The average time to complete the survey was 30 minutes, with an average compensation of \$80 for one in ten participants randomly selected for payment. Payment processing occured through online banking platform Nequi within 15 business days of participation.

@@

The CL model specifies that the probability of individual $i$ choosing alternative $j$ from choice set $C_i$ is given by:

@@

\section{Robustness check}

>>>>

\usepackage{blindtext}
\usepackage[colorlinks=true, citecolor=blue, linkcolor=blue, urlcolor=blue,breaklinks]{hyperref}

% \pdfcompresslevel=0
% \pdfobjcompresslevel=0

@@

\section*{Abstract}
Economic connectivity, defined as the share of high-SES connections in one's network, is a strong correlate of labor market income. Yet, low-SES individuals are typically at a disadvantage when it comes to knowing the right people. Referral hiring leverages networks and make explicit the role of economic connectivity where taste-based biases could further exacerbate low-SES outcomes. We conduct a field experiment with 734 university students to study the network compositions of different SES groups. We leverage enrollment networks to identify all potential referral candidates and conduct an incentivized referral exercise to reveal SES biases within these choice sets. We find that the university enrollment networks are highly segregated, with low-SES and high-SES individuals having a higher share of same-SES connections in their networks due to program selection (12\% and 31\% respectively). When considering ex post actualized choice sets for the observed referrals, the segregation becomes worse: Low-SES individuals connect with other low-SES individuals at rates 30\% higher than the university share, while high-SES individuals connect with other high-SES individuals at rates 55\% higher than the university share. Yet, we find no bias against low-SES individuals once we account for network structures. We randomly assign half of the participants to a condition where their referral candidate receives a fixed bonus on top of pay-for-performance referral incentives. We find that additional incentives for the referral candidate do not change social proximity with the referral nor the referral quality. Our findings suggest that systematic segregation patterns in networks that alter choice sets matter more than taste-based SES biases in referrals, and highlight the potential for institutional action in promoting SES diversity. \medskip \\
\textbf{JEL Classification:} C93, J71, D85, Z13 \\
\textbf{Keywords:} social capital, social networks, referral hiring, socioeconomic status, field experiment

\pagebreak
\section{Introduction}
Equally qualified individuals in terms of productivity face different labor market outcomes based on their socioeconomic status \citep{stansbury2024class}. This persistent inequality undermines meritocratic ideals and represents a substantial barrier to economic mobility. A key driver of SES-based inequality in the labor market stems from differences in social capital.\footnote{See for example \citet{loury_dynamic_1977,bourdieu_forms_1986} for pioneering work on the relationship between social position and human capital acquisition.} Economic connectivity, defined as the share of high-SES connections among low-SES individuals, is the most important facet of social capital because it correlates strongly with labor market income \citep{chetty2022na}. In this sense, a lack of social capital means lack of access to individuals with influential (higher paid) jobs and job opportunities. It implies having worse outcomes when using one's network to find jobs conditional on the capacity to leverage one's social network.\footnote{See for example \citet{lin1981asr,mouw2003asr} for differential outcomes while using contacts in job search, and \citet{smith2005ajs,Pedulla2019RaceAN} specifically for the effects of race conditional on network use.}

Referral hiring—the formal or informal process where firms ask workers to recommend qualified candidates for job opportunities—is a common labor market practice that makes differences in social capital evident.\footnote{Referrals solve some frictions in the search and matching process and benefit both job-seekers and employers. As a consequence, referral candidates get hired more often, have lower turnover, and earn higher wages \citep{brown2016informal,dustmann2016referral,friebel2023employee}.} Since referrals originate from the networks of referrers, the composition of referrer networks becomes a crucial channel that propagates inequality. Similar individuals across socio-demographic characteristics form connections at higher rates \citep{mcpherson2001birds}, making across-SES (low-to-high) connections less likely than same-SES connections \citep{chetty2022na}. Referrals will thus reflect similarities in socio-demographic characteristics present in networks even in the absence of biases in the referral procedure—that is, even when referring randomly from one's network according to some productivity criteria.

Yet, experimental evidence shows referrals can be biased even under substantial pay-for-performance incentives beyond what is attributable to differences in network compositions, at least in the case of gender \citep{beaman_job_2018,hederos2025gender}. A similar bias against low-SES individuals may further exacerbate their outcomes. If job information is in the hands of a select few high-SES individuals to whom low-SES individuals already have limited network access due to their lack of economic connectivity, and high-SES referrers are biased against low-SES individuals—referring other high-SES individuals at higher rates than their network composition would suggest—we should expect referral hiring to further disadvantage low-SES individuals.

The empirical question we answer in this paper is whether referrers exhibit bias against low-SES peers after accounting for differences in network SES composition. We also evaluate the causal impact of two different incentive structures on referral behavior.

In this study, we examine inequalities related to SES by curating a university-wide network dataset comprising over 4,500 students for whom classroom interactions are recorded along with individual attributes. We focus on the role of SES in referrals by experimentally investigating whether individuals who are asked to refer a peer tend to refer a same-SES candidate. We also explore potential mechanisms behind referral patterns by randomizing participants into two different incentive structures. To this end, we conducted a lab-in-the-field experiment with 734 students at a Colombian university. We instructed participants to refer a qualified student for tasks similar to the math and reading parts of the national university entry exam (equivalent to the SAT in the US system). To incentivize participants to refer qualified candidates during the experiment, we set earnings to depend on referred candidates' actual university entry exam scores.

Referral hiring in the labor market can range from firm-level formal referral programs asking employees to bring candidates to simply passing on job opportunities between network members \citep{topa2019social}. Since our participants are students at the university and refer based on exam scores, we abstract away from formal referral programs with defined job openings. Our setting instead resembles situations where contacts share opportunities with each other without requiring the referred candidate to take any action and without revealing the referrer's identity. This eliminates reputational concerns since there is no hiring employer. It also establishes a lower bound on the expected reciprocity for the referrer when combined with pay-for-performance incentives \citep{witte2021workers,bandiera2009social}. At the same time, referring based on university entry exam scores is still an objective, widely accepted measure of ability. We show evidence that referrers in our setting not only possess accurate information about these signals but can also screen more productive individuals from their university network.

In a university setting, class attendance provides essential opportunities for face-to-face interaction between students. This is a powerful force that reduces network segregation by providing ample opportunities to meet across SES groups, because of exposure to an equal or higher level of high-SES individuals compared to the general population \citep{chetty2022n}.\footnote{In a different sample from the same university population, \citet{diaz_peer_2025} show this holds true for the highest-SES individuals at this institution, accounting for about 6\% of their sample but less than 5\% of Colombian high-school graduates \citep{fergusson2021desigualdad}.} The very high level of income inequality in Colombia makes SES differences extremely visible in access to tertiary education, where rich and poor typically select into different institutions \citep{jaramillo-echeverri2023}. However, in the particular institutional setting we have chosen for this study, different SES groups mix at this university, allowing us to focus on SES diversity within the institution. At the same time, as students take more classes together, their similarities across all observable characteristics tend to increase \citep{kossinets_origins_2009}. This is an opposite force that drives high- and low-SES networks to segregate. We observe the net effect of these two opposing forces using administrative data and construct class attendance (enrollment) networks for 734 participants based on the number of common courses they have taken together with other students. This allows us to directly identify aggregate characterizations of different SES groups' network compositions as a function of courses taken (e.g., in same-SES share), as well as the individual characteristics of network members who receive referrals among all possible candidates.

We find strong evidence that networks of high- and low-SES participants exhibit same-SES bias. On average, both groups connect with their own SES group at higher rates than would occur randomly given actual group shares at the university (12\% for low-SES and 31\% for high-SES). As students take more courses together within the same program, their networks dwindle in size and become even more homogeneous in SES shares. At 12 courses together (the median number of courses taken together among referrals), the same-SES share increases to 30\% above the university share for low-SES students and 55\% above for high-SES students. We identify selection into academic programs as a key mechanism explaining this phenomenon: The private university where our study took place implements exogenous cost-based program pricing and does not offer SES-based price reductions. This results in programs with very large cost differences within the same university, with some programs costing up to six times the cheapest one. We find that the average yearly fee paid per student increases with SES, and the high-SES share in the most expensive program at the university—medicine—drives a large part of the network segregation across SES groups.

Do segregated networks account for the differences in SES referral rates across SES groups? Same-SES referrals are 17\% more common than referrer networks suggest. Controlling for differences in network compositions, we find that the entirety of the bias is driven by low-SES referrers. We find no bias against low-SES peers beyond what is attributable to differences in network composition. Regardless of SES, participants refer productive individuals, and referred candidates are characterized by a very high number of courses taken together. The latter underlies the impact of program selection on the intensity of social interaction, where participants activate smaller and more homogeneous parts of their networks for making referrals. Our treatment randomized participants across two different incentive schemes by adding a substantial monetary bonus (\$25) for the referred candidate on top of the pay-for-performance incentives. We provide evidence that treatment incentives did not change referral behavior across the same-SES referral rate, the number of courses taken together with the referral candidate, and the candidate's exam scores. We interpret the lack of differences in the number of courses taken together as further evidence that referrals go to strong social ties across both treatments regardless of the incentive structure.\footnote{This follows directly from earlier evidence showing that referrals tend to go to strong ties, i.e., close friends and/or family members \citep{gee2017social,kramarz2014strong,wang2013marriage}.}

Our main empirical contribution to the experimental referral literature is our observation of the entire network that characterizes the referral choice set. Earlier research compares referrals made across different incentive structures and makes inferences about the counterfactual. For example, \citet{beaman_who_2012} compared referrers paid based on their referred candidate's productivity instead of receiving a fixed finder's fee, and \citet{beaman_job_2018} compared referrers who were restricted to refer either a male or female candidate instead of choosing freely. While \citet{pallais_why_2016} recruited a random sample of non-referred workers for comparison with referred ones, none of the previous studies could identify the entire referral choice set and provide a direct comparison to those who were referred by the participants. Observing the entire network allows us to identify biases in referrals in a more natural way, without imposing restrictions on the choice sets. A similar approach to ours is \citet{hederos2025gender}, who elicited friendship networks by asking referrers to name 5 close friends. Their findings suggest only half of those who were referred were from the elicited friendship network, and thus represent an incomplete observation of the entire referral choice set. We take our analysis one step further by requesting referrals from the enrollment network, where we have complete information on every single connection that may or may not receive a referral. This allows us to neatly separate the effect of network composition from any potential biases stemming from the referral procedure itself.

Second, we build upon the earlier work on inequalities in referrals and the role of SES differences. The reliance of labor markets on referrals, coupled with homophily in social networks, can lead to persistent inequalities in wages and employment \citep{montgomery_social_1991,calvo2004effects,bolte_role_2021}. The premise of these models is that referrals exhibit homophily, so that employees are more likely to refer workers of their own race, gender, SES, etc. Supporting evidence shows that low-SES individuals have networks with lower shares of high-SES individuals, which partly explains why they have worse labor market outcomes \citep{stansbury2024class,chetty2022na}. We contribute by separately identifying the role of network homophily (the tendency to connect with similar others) and referral homophily (the tendency to refer similar others). Our results suggest that network homophily, rather than referral homophily, drives SES inequality in our setting.

To our knowledge, \citet{diaz_peer_2025} are the first to study SES biases in referrals, and our study is conceptually the closest to theirs. Drawing from a similar sample at the same institution, \citet{diaz_peer_2025} focus on referrals from first-year students made within mixed-program classrooms and find no evidence for an aggregate bias against low-SES individuals. We also find no aggregate bias against low-SES individuals in referrals beyond what is attributable to differences in network structure. Our setup differs as we sample from students who completed their first year and impose no limits on referring from a classroom. This has several implications: We find that referrals in our setup go to individuals within the same program, and that programs have different SES shares which become even more accentuated as students take more courses together. While networks drive inequality in referral outcomes because of the institutional environment in our sample, we have no reason to believe first-year student networks in \citet{diaz_peer_2025} have similar levels of segregation to begin with. Our findings suggest that implementing more mixed-program courses that allow for across-SES mixing should be a clear policy goal to reduce segregation \citep{alan2023troeasa,rohrer2021po}.

The remainder of the paper is organized as follows. Section \ref{s2} begins with the background and setting in Colombia. In Section \ref{s3} we present the design of the experiment. In Section \ref{s4} we describe the data and procedures. Section \ref{s5} discusses the results of the experiment and Section \ref{s6} introduces robustness checks. Section \ref{s7} concludes. The Appendix presents additional tables and figures as well as the experiment instructions.

@@

Our study takes place at UNAB, a medium-sized private university in Bucaramanga, Colombia with approximately 6,000 enrolled undergraduate students. The university's student body is remarkably diverse, with about 35\% of students classified as low-SES and 15\% as high-SES. This diversity at UNAB provides a unique research setting. Colombian society is highly unequal and generally characterized by limited interaction between social classes, with different socioeconomic groups separated by education and geographic residence.\footnote{Colombia has consistently ranked as one of the most unequal countries in Latin America \citep{barretoherrera2024regional}, with the richest decile earning 50 times more than the poorest decile \citep{comisioneconomicaparaamericalatinayel2023}. This economic disparity is reflected by a highly stratified society with significant class inequalities and limited class mobility \citep{angulo2012movilidad,garcia2015loteria}.} Despite significant financial barriers, many lower and middle-SES families prioritize university education for their children \citep[p. 103]{hudson_colombia_2010}. UNAB represents one of the few environments in Colombia where sustained inter-SES contact occurs naturally (see Figure \ref{school_quality_income}).

\begin{figure}[H]
  \centering
  \caption{Income, performance, and university choice in Colombia}\label{school_quality_income}
  \includegraphics[width=0.6\linewidth]{../figures/school_quality_income.png}
  \begin{tablenotes}
    \footnotesize
    \item[] \textit{Note:} This figure shows the average score national university entry exam by monthly family income and type of higher education institution. With average student score in the 65-70 band, the private university where we conducted this experiment caters to low- and high-SES students. Figure reproduced from \citet{fergusson2021distincion}.
  \end{tablenotes}
\end{figure}

To operationalize SES in our study, we rely on Colombia's established classification system. In 1994, Colombia introduced a nationwide system that divides the population into 6 strata based on housing characteristics and neighborhood amenities.\footnote{Initially designed for utility subsidies from higher strata (5 and 6) to support lower strata (1 to 3), it now extends to university fees and social program eligibility. Stratum 4 neither receives subsidies nor pays extra taxes. This stratification system largely aligns with and potentially reinforces existing social class divisions \citep{uribe2008estratificacion, guevara2019spatializing}.} We use this classification as our measure of SES in the experiment. Students in strata 1 to 2 are categorized as low-SES, strata 3 to 4 as middle-SES, and those in strata 5 to 6 as high-SES.

For our experiment, we invited via email all 4,417 UNAB undergraduate students who had completed their first year at the university at the time of recruitment. The 837 students who joined (19\%) vary in terms of their academic programs, SES, and progress in their studies. This experimental setup provides a unique opportunity for collaborative inter-class contact on equal status, whose positive effects on reducing discrimination are causally documented \citep{rao2019familiarity, mousa2020building,lowe2021types}.

The institutional setup at UNAB sets up the social contact structure and facilitates the network analysis. Undergraduate programs at UNAB are spread across two semesters, with each individual course lasting one semester. Students take between 5 to 7 courses per semester, with programs lasting anywhere between 4 to 12 semesters (2 to 6 years). Medicine, the largest program by size at UNAB, lasts for 12 semesters, followed by engineering programs at 10 semesters. Most remaining programs last for about 8 to 10 semesters, with specialized programs for immediate entry into the workforce lasting only 4 semesters.

@@

\subsection{Productivity measures}
To establish an objective basis for referral productivity, we use national university entry exam scores (SABER 11). These scores provide pre-existing, comparable measures of ability across two domains relevant for the labor market. By using existing administrative data, we eliminate the need for additional testing and ensure that all eligible students have comparable productivity measures. The scores we use in this experiment consist of critical reading and mathematics parts.

@@

\subsection{Referral task}
Building on these productivity measures, our main experimental task involves peer recommendations. After eliciting beliefs about their own performance, participants engage in incentivized peer recommendations. For both test areas (critical reading and mathematics), participants recommend one peer they believe excels in that domain. We first present an example question from the relevant test area to clarify what skills are being assessed. Participants then type the name of their recommended peer, with the system providing autocomplete suggestions from enrolled students who have taken the test (see Figure \ref{interface}).

@@

Participants can only recommend students with whom they have taken at least one class during their university studies. This requirement ensures that referrals are based on actual peer interactions and overlaps with the enrollment network that we construct. The order in which participants make recommendations across the two areas is randomized.

We incentivize referrals using a productivity-based payment scheme that rewards participants for recommending highly ranked peers. Referrers earn increasing monetary rewards as the percentile ranking of their recommendation increases (see Figure \ref{piece_rate_by_score}). This payment structure provides strong incentives to screen for highly ranked peers, with potential earnings up to \$60 per recommendation. We multiply the piece rate coefficient associated with the percentile rank by the actual test scores of the recommendation to calculate earnings.\footnote{Due to the selection into the university, the actual test score distribution has limited variance. Below a certain threshold students cannot qualify for the institution and choose a lower ranked university, and above a certain threshold they have better options to choose from.}

@@

\subsection{Treatment variation}
To examine how different incentive structures affect referral behavior, we implement a treatment that varies payment to recommended peers. We implement a between-subjects treatment that varies whether the recommended peer also receives payment. In the \textbf{Baseline} treatment, only the referrer can earn money based on their recommendation's productivity. The \textbf{Bonus} treatment adds an additional fixed payment of \$25 to any peer who is recommended in the randomly selected area for payment. This payment is independent of the peer's actual productivity (see Figure \ref{tab:treatments}).

@@

\subsection{Belief elicitation}
To understand participants' information and expectations, we collect incentivized beliefs at key points. We elicit incentivized beliefs at two points in the experiment. First, before making referrals, participants report their beliefs about their own percentile ranking in each test area. Second, after making each referral, participants report their beliefs about their recommendation's percentile ranking. For both belief elicitation tasks, participants earn \$5 if their guess is within 7 percentiles of the true value. This tolerance level is expected to balance precision with the difficulty of the task.

@@

\section{Sample, Incentives, and Procedure} \label{s4}
We invited all 4,417 UNAB undergraduate students who had completed their first year at the university at the time of recruitment to participate in our experiment. A total of 837 students participated in the data collection, yielding a 19\% response rate. Our final sample consists of 734 individuals who referred peers with whom they had taken at least one class together, resulting in an 88\% success rate for the sample. We randomly allocated participants to either \textbf{Baseline} or \textbf{Bonus} treatments.

Table \ref{tab:balance} presents key demographic characteristics and academic performance indicators across treatments (see Appendix Table \ref{tab:selection} for selection). The sample is well-balanced between the \textbf{Baseline} and \textbf{Bonus} conditions and we observe no statistically significant differences in any of the reported variables (all \textit{p} values $> 0.1$). Our sample is characterized by a majority of middle-SES students with about one-tenth of the sample being high-SES students. The test scores and GPA distributions are balanced. On average, participants had taken 3.8 courses together with members of their network, and the average network consisted of 175 peers.

@@

The experiment was conducted online through Qualtrics, with participants recruited from active UNAB students. To ensure data quality while managing costs, we randomly selected one in ten participants for payment. Selected participants received a fixed payment of \$17 for completion, plus potential earnings from one randomly selected belief question (up to \$5) and one randomly selected recommendation question (up to \$60). This structure resulted in maximum total earnings of \$82. The average time to complete the survey was 30 minutes, with an average compensation of \$80 for the one in ten participants randomly selected for payment. Payment processing occurred through online banking platform Nequi within 15 business days of participation.

@@

The conditional logit model specifies that the probability of individual $i$ choosing alternative $j$ from choice set $C_i$ is given by:

@@

\section{Robustness check} \label{s6}
